### 2018-03-25 example.tex
Update on Overleaf.

\documentclass[letterpaper]{article}

\usepackage{natbib,alifeconf}  %% The order is important

% *****************
%  Requirements:
% *****************
%
% - All pages sized consistently at 8.5 x 11 inches (US letter size).
% - PDF length <= 8 pages for full papers, <=2 pages for extended
%    abstracts.
% - Abstract length <= 250 words.
% - No visible crop marks.
% - Images at no greater than 300 dpi, scaled at 100%.
% - Embedded open type fonts only.
% - All layers flattened.
% - No attachments.
% - All desired links active in the files.

% Note that the PDF file must not exceed 5 MB if it is to be indexed
% by Google Scholar. Additional information about Google Scholar
% can be found here:
% http://www.google.com/intl/en/scholar/inclusion.html.

% If your system does not generate letter format documents by default,
% you can use the following workflow:
% latex example
% bibtex example
% latex example ; latex example
% dvips -o example.ps -t letterSize example.dvi
% ps2pdf example.ps example.pdf

% For pdflatex users:
% The alifeconf style file loads the "graphicx" package, and
% this may lead some users of pdflatex to experience problems.
% These can be fixed by editing the alifeconf.sty file to specify:
% \usepackage[pdftex]{graphicx}
%   instead of
% \usepackage{graphicx}.
% The PDF output generated by pdflatex should match the required
% specifications and obviously the dvips and ps2pdf steps become
% unnecessary.

% Note:  Some laser printers have a serious problem printing TeX
% output. The use of ps type I fonts should avoid this problem.

\title{Something about lineage metrics}
%\author{Chris Adami$^{1}$, Ryoichi Seki$^{1,2}$ \and Robel Yirdaw$^2$ \\
%\mbox{}\\
%$^1$California Institute of Technology, Pasadena, CA 91125 \\
%$^2$California State University, Northridge, CA 91330 \\
%adami@caltech.edu} % email of corresponding author

% For several authors from the same institution use the same number to
% refer to one address.
%
% If the names do not fit well on one line use
%         Author 1, Author 2 ... \\ {\Large\bf Author n} ...\\ ...
%
% If the title and author information do not fit in the area
% allocated, place \setlength\titlebox{<new height>} after the
% \documentclass line where <new height> is 2.25in

\begin{document}
\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

Evolution is an emergent effect of events that occur on a much finer-grained temporal scale. While this is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results. In computational evolutionary systems, we theoretically have adequate data to untangle these mechanisms. In practice, though, gathering appropriate data to do so can be challenging. Moreover, because of the vast difference between the scales that the causes and effects occur on, the quantity of such data can be overwhelming. Both of these problems can be mitigated with a standardized set of diagnostic summary metrics that paint a picture of shorter-term evolutionary dynamics within a population. Here, we will introduce a suite of such metrics and provide an intuition for what they can tell us about evolution.

\section{Methods}

\subsection{Metrics}

\subsection{Interpretation}

\subsubsection{Motivation}
Evolutionary biologists often accuse each other of making up "just-so" stories - post-hoc explanations for how an observed trait or pattern could have evolved. The problem with these stories is that, in biology, there is often no way to verify that they reflect what happened in biology. Due to evolution's stochastic nature, it is easy to come up with possible ways that something could have happened once, regardless of how repeatable they are. Theoretically, we should easily be able to avoid this problem in computational evolution by carefully checking the underlying assumptions behind our hypothesized explanation.  All too often, however, we fail to drill down to the true underlying mechanism behind an observed effect.

The metrics we propose here can help provide a check on this behavior, by making information about underlying evolutionary dynamics more readily accessible. They can only have that effect, however, if we completely understand what they are telling us about the way populations under different conditions are traversing different fitness landscapes. In order for these metrics to be useful for diagnosing the behavior of an evolving population, we need to establish ground truth for what underlying evolutionary dynamics are implied by different values of the metrics. For the purposes of building this intuition on a solid foundation, we wanted to be able to visualize the full evolutionary history of each population as it traversed the fitness landscape.

To this end, we chose problems for which we could perfectly visualize the fitness landscape, and kept track of the complete evolutionary history of all members of the population. With these data, we were able to overlay successful lineages on top of the fitness landscape. Visualizing all lineages in each experimental condition in this way gave us an efficient check on our understanding of the underlying evolutionary dynamics in that condition.

\subsubsection{Visualization methods}

Displaying the complete history of an evolving population on a three-dimensional fitness landscape entails incorporating a large quantity of information into a limited space. When projected onto two dimensions, lineages can obscure parts of the fitness landscape (and each other). To mitigate this problem, we built the visualization using the A-Frame Framework. A-Frame supports building three dimensional scenes and rendering them to a variety of platforms. In the simplest case, the visualization is rendered in WebGL, allowing it to be viewed in a standard web browser. Mouse interactions such as rotation make it possible to view the visualization from all angles, and WebGL's use of the graphics card allows it to render data-rich visualizations.

A-Frame also supports rendering the page with WebVR, allowing it to be viewed using various virtual reality headsets. These platforms allow the user to explore the data in a truly three-dimensional environment. Different headsets support different amounts of interaction. For the data interpretation in this paper, we used an Oculus Rift to provide us with fine-grained control of which part of the visualization we were looking at.

\subsection{Problems}

\section{Results and Discussion}

\section{Conclusions}

\section{Acknowledgements}

\footnotesize
\bibliographystyle{apalike}
\bibliography{example} % replace by the name of your .bib file

\end{document}

### 2018-03-26 source/scripts/potHoles.py
Updated pot holes script to operate only on run lists. No need to input not_done logs.

@@ -1,4 +1,4 @@
-import argparse, os, errno
+import argparse, os, errno, copy
 
 '''
 potHoles.py
@@ -32,21 +32,21 @@ EXEC=command.sh
 def main():
     parser = argparse.ArgumentParser(description="Pot hole filler. MUST have at least 1 not_done log or run_list.")
     parser.add_argument("-d", "--data_directory", type=str, action="append", help="Target experiment directory.")
-    parser.add_argument("-l", "--not_done_log", type=str, action="append", help="Location of log of unfinished jobs.")
     parser.add_argument("-c", "--condition", type=str, action="append", help="At least one given condition string must be job name to include in resubs.")
     parser.add_argument("-r", "--run_list", type=str, action="append", help="Location of run_list that describes original experiment conditions.")
     parser.add_argument("-dest", "--destination", type=str, help="Where to put missing runs")
     parser.add_argument("-config", "--configs", type=str, help="Where to find config directory")
+    parser.add_argument("-u", "--update", type=str, help="Update that indicates run finished (final update)")
 
     args = parser.parse_args()
 
     data_directories = args.data_directory
-    not_done_logs = args.not_done_log
     filter_jobs = bool(args.condition)
     include_conditions = args.condition
     run_lists = args.run_list
     config_dir = args.configs
     dest_dir = args.destination
+    final_update = args.update
     
     def JobFilter(job_name):
         if (not filter_jobs): return True
@@ -56,15 +56,15 @@ def main():
         return False
 
     print("Data directories: {}".format(data_directories))
-    print("Not done logs: {}".format(not_done_logs))
+    # print("Not done logs: {}".format(not_done_logs))
 
     if (filter_jobs):
         print("Filtering jobs on conditions: {}".format(include_conditions))
     if (not data_directories):
         print("Must provide at least one data directory!")
         exit()
-    if (not not_done_logs and not run_lists):
-        print("Must provide at least one log of unfinished jobs or one run_list file!")
+    if (not run_lists):
+        print("Must provide at least one run_list file!")
         exit()
 
     if (run_lists):
@@ -92,8 +92,22 @@ def main():
             run_name = run.split(" ")[0]
             run_dir = "{}_{}".format(run_name, run.split(" ")[3])
[…]

### 2018-03-28 analysis/cec_python_library/extract_dominant_lineaage_variance.py
Added variance extraction script

@@ -1,0 +1,104 @@
+#!/home/emily/anaconda2/bin/python
+###############################################################################
+# Version: 1.1
+# Last modified on: 3 April, 2016 
+# Developers: Michael G. Epitropakis
+#      email: m_(DOT)_epitropakis_(AT)_lancaster_(DOT)_ac_(DOT)_uk 
+###############################################################################
+from cec2013.cec2013 import *
+import numpy as np
+import sys
+import pandas as pd
+import glob
+import os
+
+def main():
+
+	glob_pattern = sys.argv[1]
+	outfilename = sys.argv[2]
+
+	all_data = pd.DataFrame()
+
+	# print(glob_pattern, glob.glob(glob_pattern))
+
+	problem_map = {0:4, 1:5, 2:6, 6:12}
+
+	for dirname in glob.glob(glob_pattern):
+		print(dirname)
+		run_log = dirname + "/run.log"
+		filename = dirname + "/pop_5000/phylogeny_5000.csv"
+
+		if not (os.path.exists(run_log) and os.path.exists(filename)):
+			continue
+
+		local_data = {}
+
+		with open(run_log) as run_log_file:
+			for line in run_log_file:
+				if line.startswith("Doing initial"):
+					break
+				elif not line.startswith("set"):
+					continue
+				line = line.split()
+				local_data[line[1]] = line[2]
+
+		# Create function
+		f = CEC2013(problem_map[int(local_data["PROBLEM"])])
+		assert(f.get_dimension() == 2)			
+
+		df = pd.read_csv(filename, index_col="id")
+
+		df.rename(str.strip, axis="columns", inplace=True)
+		# print(df.columns.values)
+		genotypes = [i.strip("[] ").split() for i in df["info"]]
+		genotypes = map(lambda x: [float(x[0]), float(x[1])], genotypes)
+		df["fitness"] = [f.evaluate(np.array(x)) for x in genotypes]
+		genotypes = zip(*genotypes)
+
+		df["x"] = genotypes[0]
+		df["y"] = genotypes[1]
+
+		paths = []
+		ids  = []
+
+		next_gen = set([])
+		curr_id = int(df[df["num_orgs"] > 0]["fitness"].idxmax())
+
+		path = []
+		fits = []
+		while (curr_id > 1):
+			# print(df.loc[curr_id, :])
[…]

### 2018-03-28 source/scripts/potHoles.py
Fixed box in potholes script.

@@ -103,7 +103,7 @@ def main():
             # If we can find it, did it finish?
             if (found):
                 contents = None
-                with open(os.path.join(data_directory, run_dir), "r") as fp:
+                with open(os.path.join(data_directory, run_dir, "run.log"), "r") as fp:
                     contents = fp.read()
                 if not "Update: {} Max score:".format(final_update) in contents:
                     found = False

### 2018-03-28 source/scripts/potHoles.py
Added a missing runs log to pot holes script.

@@ -88,6 +88,7 @@ def main():
         print("Total number of runs: " + str(len(all_runs)))
         # 2) Try to find run
         missing_runs = {}
+        old_run_log = []
         for run in all_runs:
             run_name = run.split(" ")[0]
             run_dir = "{}_{}".format(run_name, run.split(" ")[3])
@@ -113,8 +114,12 @@ def main():
                 missing_runs[run_dir] = {}
                 missing_runs[run_dir]["command"] = " ".join(run.split(" ")[1:])
                 missing_runs[run_dir]["run_dir"] = os.path.join("${DEST_DIR}", run_dir)
+                old_run_log.append(os.path.join(data_directory, run_dir))
         
         print("Total number of missing runs: " + str(len(missing_runs)))
+
+        with open("missing_runs.log", "w") as fp:
+            fp.write("\n".join(old_run_log))
         
         # 3) Build Qsub file. 
         qsub = qsub_base.replace("[[JOB_CONFIG:-t]]", "{}-{}".format(1,len(missing_runs)))

### 2018-03-28 paper.tex
Update on Overleaf.

--- 2018-03-25 example.tex
@@
 
-Evolution is an emergent effect of events that occur on a much finer-grained temporal scale.
+Evolution is an emergent effect of events (\textit{e.g.} replication, variation, and competition) that occur on a much finer-grained temporal scale.
 While this is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results.
@@
 Both of these problems can be mitigated with a standardized set of diagnostic summary metrics that paint a picture of shorter-term evolutionary dynamics within a population.
-Here, we will introduce a suite of such metrics and provide an intuition for what they can tell us about evolution.
+Here, we  introduce a suite of such metrics and provide an intuition for what they can tell us about evolution.
+% We test our intuitions on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths.
 
-\section{Methods}
+% Talk about abstract idea of a lineage being a sequence of states
+% A lineage can be operationally defined as a sequence of states
+% talk about different things those states could be
 
-\subsection{Metrics}
+%  Talk about metrics being a useful concept for abstracting sequence of states (because doing stats on sequences of states is hard)
 
-\subsection{Interpretation}
+%Lit review - people have made pictures but it's hard to draw abstract large-scale inferences from them
 
-\subsubsection{Motivation}
-Evolutionary biologists often accuse each other of making up "just-so" stories - post-hoc explanations for how an observed trait or pattern could have evolved.
+% Introduce metrics
+
+For the most part, the metrics we propose here will center around the evolutionary history of lineages in the population.
+
+% @AML: Currently taking a crack at this bit of the intro (introducing lineages, etc) on paper at the moment!
+
+% Something about: we often look at fitness over a lineage (i.e. fitness over time), we can look at other things...
+
+\section{Metrics}
+
+\subsection{Lineage Length}
+\subsection{Mutation Accumulation} 
+% Count/Magnitude (by type, aggregated etc
+\subsection{Phenotypic Volatility}
+% Useful things that this metric could identify:
+% - Mutational bet hedging (h
+% - Plasticity (high volatility)
+% - Neutral landscapes (expect low volatility)
+
+\subsection{Summary statistics}
+
+\section{Test Problems}
+To understand the metrics defined above, the test problems used need to be well understood and studied.
+The benchmark functions from [URL OR TITLE?] meet both of these requirements and allow us to visualize the actual fitness landscape, due to the low dimensionality of the problems.
+For each problem, the X and Y coordinates offered by a given organism are translated by the function into a fitness value.
+We chose a diverse subset of these functions (Himmelblau, Shubert, Composition Function 2, and Six-Humped Camel Back) as our test problems in order to gain a broad understanding of our metrics.
+
+\section{Data Analysis}
+
+% Trim dramatically
+%Evolutionary biologists often accuse each other of making up "just-so" stories - post-hoc explanations for how an observed trait or pattern could have evolved.
 The problem with these stories is that, in biology, there is often no way to verify that they reflect what happened in biology.
@@
 
-The metrics we propose here can help provide a check on this behavior, by making information about underlying evolutionary dynamics more readily accessible.
+%The metrics we propose here can help provide a check on this behavior, by making information about underlying evolutionary dynamics more readily accessible.
 They can only have that effect, however, if we completely understand what they are telling us about the way populations under different conditions are traversing different fitness landscapes.
@@
 
-To this end, we chose problems for which we could perfectly visualize the fitness landscape, and kept track of the complete evolutionary history of all members of the population.
+%To this end, we chose problems for which we could perfectly visualize the fitness landscape, and kept track of the complete evolutionary history of all members of the population.
 With these data, we were able to overlay successful lineages on top of the fitness landscape.
 Visualizing all lineages in each experimental condition in this way gave us an efficient check on our understanding of the underlying evolutionary dynamics in that condition.
-
-\subsubsection{Visualization methods}
 
@@
 
-\subsection{Problems}
+\section{Results and Discussion}
+% Changing mutation: here's what happens when we change mutation on each problem
+% - Metric 1
+% Problem 1
+% Problem 2
+% Problem 3
+% Problem 4
+% - Metric 2
+% ...
 
-\section{Results and Discussion}
+% Changing selection
+% - Metric 1
+% Problem 1
+% Problem 2
+% Problem 3
+% Problem 4
+% - Metric 2
 
@@
 
+\section{Text Bank Because I'm not totally sure where things should go}
+% Some gibberish on lineages
+A lineage describes a chain of ancestors and their descendants where links in the chain are (typically) connected via parent-offspring relationships.
+In nature, every extant individual has a unique lineage that can be traced back to a single common ancestor.
+Depending on the modes of replication and variation employed along the lineage (\textit{e.g.} sexual reproduction, horizontal gene transfer), a lineage can be more or less challenging to disentangle.
+% something something something, we focus on asexual populations w/no HGT because easy starting place.
+
+% Some gibberish on how we can represent lineages as sequences of states
+Lineages can be imagined as a sequence of states where each state represents an individual.
+We can characterize each of these states: a phenotype, a genotype, a set of mutations (which may be empty) that differentiate genotype from parent genotype, time of birth/death, etc.
+Given sequences, can apply diagnostic metrics...
+
+% Some gibberish on abstractions over the sequences of states
+We can start to make abstractions: state = genotype, state = phenotype, state = location in environment.
+Lineage length is measured over these abstractions:
+%- Length in number of individuals 
+%- Length in number of genotypes
+%- Length in number of phenotypes
+
 \end{document}

### 2018-03-28 analysis/data_analysis.Rmd
Added figs

---
title: "Lineage Metrics Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(readr)

```

```{r, echo=FALSE}
alex_real_value_data <- read_csv("alex_real_value_data.csv")
emily_real_value_data <- read_csv("real_value_data.csv")
real_value_data <- merge(alex_real_value_data, emily_real_value_data, all = TRUE)
real_value_data$problem <- sapply(as.character(real_value_data$p),switch,"0"="Himmelblau","1"="Six-hump camel back","2"="Shubert", "6"="Composition function 2")
real_value_data$problem <- as.factor(real_value_data$problem)
real_value_data$selection <- sapply(as.character(real_value_data$selection),switch,"0"="Tournament","1"="Lexicase","2"="Eco-EA", "3"="MAP-Elites", "4"="Roulette", "5"="Drift")
real_value_data$selection <- as.factor(real_value_data$selection)
real_value_data$seed <- as.factor(real_value_data$seed)
real_value_data$ts <- as.factor(real_value_data$ts)
real_value_data$elite <- as.factor(real_value_data$elite)
real_value_data$mutation_rate <- as.factor(real_value_data$mutation_rate)
tournament_data <- subset(real_value_data, real_value_data$selection == "Tournament" & real_value_data$elite == 0)
```

```{r}
#see how many finished
summary(subset(real_value_data, real_value_data$update == 5000))

```

First, let's make sure there's nothing weird going on with fitness:

```{r, echo=FALSE}
ggplot(tournament_data) + stat_summary(aes(x=update, y=max_fitness, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
ggsave("../figs/all_tournament_fitness.png")
```

The most important thing to observe from that is probably that mutation rate seems to have a notable effect on the fitness dynamics of the Shubert function.

Onto the metrics.

## Phenotypic volatility

```{r, echo=FALSE}
ggplot(tournament_data) + stat_summary(aes(x=update, y=mean_phenotypic_volatility, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
ggsave("../figs/all_phenotypic_volatility.png")
```

Oh right, that one kind of breaks in these problems...

## Phenotype count

```{r, echo=FALSE}
ggplot(tournament_data) + stat_summary(aes(x=update, y=mean_unique_phenotypes, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
ggsave("../figs/all_phenotype_count.png")

```

These too.

## Deleterious steps

```{r, echo=FALSE}
ggplot(tournament_data) + stat_summary(aes(x=update, y=mean_deleterious_steps, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
ggsave("../figs/all_deleterious_steps.png")

```

In general, weaker selection allows more deleterious steps, as does higher mutation rate.

```{r, echo=FALSE}
ggplot(tournament_data) + stat_summary(aes(x=update, y=dominant_deleterious_steps, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
ggsave("../figs/dominant_deleterious_steps.png")

```

Looks pretty similar.

```{r, echo=FALSE}
ggplot(subset(tournament_data, tournament_data$update==5000)) + geom_point(aes(x=max_deleterious_steps, y=dominant_deleterious_steps, color=ts)) + facet_wrap(~problem*mutation_rate, ncol=8, scales="free") + theme_classic()
```

The relationship between max deleterious steps and dominant deleterious steps stops being a straight line at higher mutation rates.

# Graphs for paper

## Varying mutation rate:

### Deleterious steps:

```{r, echo=FALSE}
ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=max_pairwise_distance, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
```

```{r, echo=FALSE}
ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=dominant_deleterious_steps, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
ggsave("../figs/dom_deleterious_mutation_rate.png")
```

There's possibly an interesting story to tell about variance:

```{r, echo=FALSE}
ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=variance_deleterious_steps, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
ggsave("../figs/variance_deleterious_mutation_rate.png")
```

### Mutation magnitude

```{r, echo=FALSE}
ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=mean_normal_mutations_on_lineage, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
```

Presumably the reason that .1 is ludircously high in everything except camel-back is that on the other landscapes it's jumping between peaks. Let's look at it without .1:

```{r, echo=FALSE}
ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4 & as.numeric(tournament_data$mutation_rate) < 8)) + geom_boxplot(aes(x=mutation_rate, y=mean_normal_mutations_on_lineage, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
```

### 2018-03-28 paper.tex
Update on Overleaf.

@@
 
-For the most part, the metrics we propose here will center around the evolutionary history of lineages in the population.
+%For the most part, the metrics we propose here will center around the evolutionary history of lineages in the population.
 
 % @AML: Currently taking a crack at this bit of the intro (introducing lineages, etc) on paper at the moment! 
+A lineage describes a continuous line of descent, linking ancestors and their descendants.
+A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
+Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for
 
@@
 % Changing mutation: here's what happens when we change mutation on each problem
+\begin{figure}
+\includegraphics[width=3.5in]{figs/dom_deleterious_mutation_rate.png}
+\end{figure}
+
 % - Metric 1
@@
 
-\section{Text Bank Because I'm not totally sure where things should go}
-% Some gibberish on lineages
-A lineage describes a chain of ancestors and their descendants where links in the chain are (typically) connected via parent-offspring relationships.
+% \section{Text Bank Because I'm not totally sure where things should go}
+% % Some gibberish on lineages
+% A lineage describes a chain of ancestors and their descendants where links in the chain are (typically) connected via parent-offspring relationships.
 In nature, every extant individual has a unique lineage that can be traced back to a single common ancestor.
 Depending on the modes of replication and variation employed along the lineage (\textit{e.g.} sexual reproduction, horizontal gene transfer), a lineage can be more or less challenging to disentangle.
-% something something something, we focus on asexual populations w/no HGT because easy starting place.
+% % something something something, we focus on asexual populations w/no HGT because easy starting place.
 
-% Some gibberish on how we can represent lineages as sequences of states
-Lineages can be imagined as a sequence of states where each state represents an individual.
+% % Some gibberish on how we can represent lineages as sequences of states
+% Lineages can be imagined as a sequence of states where each state represents an individual.
 We can characterize each of these states: a phenotype, a genotype, a set of mutations (which may be empty) that differentiate genotype from parent genotype, time of birth/death, etc.
-Given sequences, can apply diagnostic metrics...
+% Given sequences, can apply diagnostic metrics...
 
-% Some gibberish on abstractions over the sequences of states
-We can start to make abstractions: state = genotype, state = phenotype, state = location in environment.
-Lineage length is measured over these abstractions:
-%- Length in number of individuals 
-%- Length in number of genotypes
-%- Length in number of phenotypes
+% % Some gibberish on abstractions over the sequences of states
+% We can start to make abstractions: state = genotype, state = phenotype, state = location in environment.
+% Lineage length is measured over these abstractions:
+% %- Length in number of individuals 
+% %- Length in number of genotypes
+% %- Length in number of phenotypes
 

### 2018-03-28 paper.tex
Update on Overleaf.

@@
 A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
-Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for
+Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained
 

### 2018-03-28 analysis/data_analysis.Rmd
Make figure size reasonable

@@
 ```{r, echo=FALSE}
-ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=max_pairwise_distance, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
+ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=max_pairwise_distance, color=mutation_rate)) + facet_wrap(~problem, scales="free") + theme_classic()
 ```
@@
 ```{r, echo=FALSE}
-ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=dominant_deleterious_steps, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
-ggsave("../figs/dom_deleterious_mutation_rate.png")
+ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=dominant_deleterious_steps, color=mutation_rate)) + facet_wrap(~problem, scales="free") + theme_classic()
+ggsave("../figs/dom_deleterious_mutation_rate.png", width = 3.5)
 ```
@@
 ```{r, echo=FALSE}
-ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=variance_deleterious_steps, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
-ggsave("../figs/variance_deleterious_mutation_rate.png")
+ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=variance_deleterious_steps, color=mutation_rate)) + facet_wrap(~problem, scales="free") + theme_classic()
+ggsave("../figs/variance_deleterious_mutation_rate.png", width = 3.5)
 ```

### 2018-03-28 paper.tex

--- 2018-03-28 paper.tex
@@
 A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
-Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for
+Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at
 

### 2018-03-28 paper.tex
Update on Overleaf.

--- 2018-03-28 paper.tex
@@
 A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
-Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for
+Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in moder
 

### 2018-03-28 paper.tex
Update on Overleaf.

--- 2018-03-28 paper.tex
@@
 A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
-Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained
+Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in modern web lab experiments
 

### 2018-03-28 paper.tex
Update on Overleaf.

--- 2018-03-28 paper.tex
@@
 A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
-Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at
+Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in modern web lab experiments, allowing us to replay the tape of life in exacting detail to tease apart the evolutionary recipe for whatever complex feature we happen to be interested in.
+% Yet, tracking the full details of a single lineage, much less a population of lineages, can be computational expense
 

### 2018-04-01 paper.tex

--- 2018-03-28 paper.tex
@@
 Both of these problems can be mitigated with a standardized set of diagnostic summary metrics that paint a picture of shorter-term evolutionary dynamics within a population.
-Here, we  introduce a suite of such metrics and provide an intuition for what they can tell us about evolution.
-% We test our intuitions on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths.
+Here, we introduce a suite of such metrics that operate on lineages and provide an intuition for what they can tell us about evolution.
 
-% Talk about abstract idea of a lineage being a sequence of states
-% A lineage can be operationally defined as a sequence of states
-% talk about different things those states could be
+% Useful suites of metrics in ALIFE?
+% - Activity statistics
+% - ...
 
-%  Talk about metrics being a useful concept for abstracting sequence of states (because doing stats on sequences of states is hard)
-
-%Lit review - people have made pictures but it's hard to draw abstract large-scale inferences from them
-
-% Introduce metrics
-
-%For the most part, the metrics we propose here will center around the evolutionary history of lineages in the population.
-
-% @AML: Currently taking a crack at this bit of the intro (introducing lineages, etc) on paper at the moment! 
 A lineage describes a continuous line of descent, linking ancestors and their descendants.
 A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
-Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in moder
+Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in modern wet lab experiments, allowing us to replay the tape of life in exacting detail to tease apart the evolutionary recipe for whatever complex feature we happen to be interested in.
+Yet, tracking the full details of a single lineage, much less a population of lineages, can be computationally expensive and will inevitably generate an unwieldy amount of data that can be challenging to visualize and interpret [CITATIONS].
+Summary statistics can help alleviate these issues by focusing on general trends(?) for a population rather than for each individual.
+% @AML: Need a transition from here to next paragraph about how diagnostic/summary statistics can help
+% @SPJ: Something like this?
 
-% Something about: we often look at fitness over a lineage (i.e. fitness over time), we can look at other things...
+Here, we demonstrate three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility.
+For each, we discuss their application and our intuitions for what they can tell us about evolution.
+We evaluate our intuitions on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths.
+% @AML: we'll want to say something here about how we talk about how these metrics extend to genetically discrete organisms (e.g. GP)
 
 \section{Metrics}
+% Outlining some points to (maybe) make here; probably too bogged down in details; not even sure we want to talk about lineages as graphs:
+% - Apply metrics in context of asexual populations
+% - Establish a framework from which to think about metrics (not novel, but rarely formalized)
+% 	 - We treat lineages as sequences of states where each state represents an individual and transitions between states represent parent-offspring relationships.
+%      - More formally, lineages are directed acyclic graphs where internal nodes have in-degree = 1 (parent), out-degree = 1 (offspring).
+Original ancestor (in digital evolution, often seed of population) w/out-degree = 1.
+Extant individual w/in-degree = 1.
+%      - Transitions specify relevant variation introduced into offspring (e.g. mutations)
+%      - States specify relevant genotype and phenotype information (e.g. genome, location in space, behavior, etc) -- anything we might want to measure over a lineage
+%      - Becomes more complicated with sexual populations, but we'll stick with asexual assumption for this. 
+%  - Already, framework affords certain questions:
+%     - How many individuals are along the lineage? 
+%     - Total # and magnitude of mutations along lineage?
+%     - Distribution of mutation types (ins vs. sub/beneficial vs. deleterious)?
+%  - Further abstractions:
+%      - Sequence of genotypes along lineage (where each genotype may represent multiple individuals)
+%      - Sequence of phenotypes along lineage (where each phenotype may represent multiple individuals)
+%      - Still further abstraction: sequence of a particular aspect of genotype (e.g. genetic marker)/phenotype (e.g. location) along lineage
+%  - We discuss three classes of lineage metrics within this framework: lineage length, mutation accumulation, and phenotypic volatility. 
+%  - For each, we discuss our intuition for what they can tell us (alone and in combination) about evolution.
 
 \subsection{Lineage Length}
+% Define/explanation
+% 	- Within context of framework, lineage length is simply measure of number of states in the graph/sequence.
+%   - What this means will depend on what states represent?
+%       - number of individuals along lineage 
+%             - More interesting in experiments where generations are not dictated by system, but rather by life history strategies of organisms
+%             - Give you a diagnostic for replication rate (r vs. k strategy)
+%      - number of genotypes along lineage
+%            - length = 1: no mutations
+%            - length = number of individuals: mutation in every replication
+%            - Intuitively, should be higher at high mutation rates, lower at low mutation rates. 
+%      - number of phenotypes along lineage
+%            - low: evidence for static environment
+%            - high: might be indicative of a changing environment or ecological interactions
+%            - if number of genotypes is high, but number of phenotypes low, probably in a neutral fitness landscape
+%      - various subsets of genotype/phenotype
+%           - location in space (environment type): useful for looking at lineage through heterogeneous landscape
+%           - change in particular phenotypic trait
+%           - change in genetic markers
+%           - Distributions of above two lengths can tell you useful things about drift, selection, etc.
+
+% Here, we...
+
 \subsection{Mutation Accumulation}
-% Count/Magnitude (by type, aggregated etc
+Mutation Accumulation defines a set of measurements that track mutational changes across a lineage.
+These changes can be measured in a variety of ways, such as the magnitude of the change (for real-valued genomes), or the total count of changes (for discrete-valued genomes).
+The type of a mutation can also be tracked, and can provide insight into the distribution of mutation types for a given lineage.
+In conjunction with collected fitness information, the class of a mutation (\textit{e.g.} beneficial, deleterious, or neutral) can also be tracked.
+% @SPJ: TODO
+% Mutation accumulation over a lineage 
+%  - Measured in a variety of ways
+%   - For real-valued genomes, we can track magnitudes of changes
+%   - For discrete-valued genomes, we can track counts of changes
+%      - Track by type to get distribution of mutation types
+%   - In conjunction with fitness information, can track classes of mutation: beneficial, deleterious, neutral
+%      - Again, can extract a distribution; different environments/experiments, you might expect different distributions
+%      - Importance of deleterious steps; valley crossing, etc
+
+% Here, we...
+
 \subsection{Phenotypic Volatility}
+% Define/describe
+% .... 
 % Useful things that this metric could identify:
-% - Mutational bet hedging (h
+% - Mutational bet hedging (high volatility)
 % - Plasticity (high volatility)
@@
 
+% Here, we...
+
 \subsection{Summary statistics}
+
+Each of these metrics can be calculated for each member of the population at each time step.
+Doing so, however, would produce an amount of data so large that it would be difficult to make sense of.
+Instead, we need to come up with ways to generate useful summaries.
+There are two main approaches to doing so: 1) choose a small number of representative lineages from a given time point, 2) collect summary statistics about the distribution of metric values across the population.
 
@@
 To understand the metrics defined above, the test problems used need to be well understood and studied.
-The benchmark functions from [URL OR TITLE?] meet both of these requirements and allow us to visualize the actual fitness landscape, due to the low dimensionality of the problems.
+The benchmark functions from the GECCO Competition on Niching Methods meet both of these requirements and allow us to visualize the actual fitness landscape, due to the low dimensionality of the problems ~\citep{li_benchmark_2013}.
 For each problem, the X and Y coordinates offered by a given organism are translated by the function into a fitness value.
 We chose a diverse subset of these functions (Himmelblau, Shubert, Composition Function 2, and Six-Humped Camel Back) as our test problems in order to gain a broad understanding of our metrics.
+We used the implementations of these problems at https://github.com/mikeagn/CEC2013/ (C++ for fitness calculations during evolution, Python for post-hoc analysis).
+
+\section{Implementation}
 
@@
 Visualizing all lineages in each experimental condition in this way gave us an efficient check on our understanding of the underlying evolutionary dynamics in that condition.
+In order to make these metrics useful in the long run, it is critical that we have an accurate understanding of how various measurements correspond to the actual behavior of lineages.
+The most direct way to confirm our expectations is to visualize the path that each lineage takes across the fitness landscape, mapping the x, y, and z (fitness) coordinates of each ancestor of each member of the population.
+Creating such a visualization entails incorporating a large quantity of information into a limited space.
+When projected onto two dimensions, lineages can obscure parts of the fitness landscape (and each other).
+To mitigate this problem, we built a thee-dimesnional visualization using the A-Frame Framework [cite].
 
-Displaying the complete history of an evolving population on a three-dimensional fitness landscape entails incorporating a large quantity of information into a limited space.
-When projected onto two dimensions, lineages can obscure parts of the fitness landscape (and each other).
-To mitigate this problem, we built the visualization using the A-Frame Framework.
 A-Frame supports building three dimensional scenes and rendering them to a variety of platforms.
@@
 Mouse interactions such as rotation make it possible to view the visualization from all angles, and WebGL's use of the graphics card allows it to render data-rich visualizations.
-
 A-Frame also supports rendering the page with WebVR, allowing it to be viewed using various virtual reality headsets.
@@
 \includegraphics[width=3.5in]{figs/dom_deleterious_mutation_rate.png}
+\end{figure}
+\begin{figure}
+\includegraphics[width=3.5in]{figs/variance_deleterious_mutation_rate.png}
 \end{figure}
@@
 \bibliographystyle{apalike}
-\bibliography{example} % replace by the name of your .bib file
-
-% \section{Text Bank Because I'm not totally sure where things should go}
-% % Some gibberish on lineages
-% A lineage describes a chain of ancestors and their descendants where links in the chain are (typically) connected via parent-offspring relationships.
-In nature, every extant individual has a unique lineage that can be traced back to a single common ancestor.
-Depending on the modes of replication and variation employed along the lineage (\textit{e.g.} sexual reproduction, horizontal gene transfer), a lineage can be more or less challenging to disentangle.
-% % something something something, we focus on asexual populations w/no HGT because easy starting place.
-
-% % Some gibberish on how we can represent lineages as sequences of states
-% Lineages can be imagined as a sequence of states where each state represents an individual.
-We can characterize each of these states: a phenotype, a genotype, a set of mutations (which may be empty) that differentiate genotype from parent genotype, time of birth/death, etc.
-% Given sequences, can apply diagnostic metrics...
-
-% % Some gibberish on abstractions over the sequences of states
-% We can start to make abstractions: state = genotype, state = phenotype, state = location in environment.
-% Lineage length is measured over these abstractions:
-% %- Length in number of individuals 
-% %- Length in number of genotypes
-% %- Length in number of phenotypes
+\bibliography{Zotero} % replace by the name of your .bib file
 

### 2018-04-02 analysis/cec_python_library/extract_dominant_lineaage_variance.py
Also track neutral and beneficial mutations

@@ -66,12 +66,29 @@ def main():
 
 		path = []
 		fits = []
+		x_magnitude = 0
+		y_magnitude = 0
+		total_magnitude = 0
+		deleterious_steps = 0
+		ben_steps = 0
+		neutral_steps = 0
 		while (curr_id > 1):
 			# print(df.loc[curr_id, :])
 			fits.append(df.loc[curr_id, "fitness"])
+			
 			path.append(" ".join(["{:.2f}".format(i) for i in [df.loc[curr_id, "x"], df.loc[curr_id, "y"], df.loc[curr_id, "fitness"]]]))
 			if (df.loc[curr_id, "parent_id"] == 1):
 				break
+			x_magnitude += abs(df.loc[df.loc[curr_id, "parent_id"], "x"] - df.loc[curr_id, "x"]) 
+			y_magnitude += abs(df.loc[df.loc[curr_id, "parent_id"], "y"] - df.loc[curr_id, "y"])
+			total_magnitude += abs(df.loc[df.loc[curr_id, "parent_id"], "x"] - df.loc[curr_id, "x"]) + abs(df.loc[df.loc[curr_id, "parent_id"], "y"] - df.loc[curr_id, "y"])
+			if (df.loc[curr_id, "fitness"] > df.loc[df.loc[curr_id, "parent_id"], "fitness"]):
+				ben_steps += 1
+			elif (df.loc[curr_id, "fitness"] == df.loc[df.loc[curr_id, "parent_id"], "fitness"]):
+				neutral_steps += 1
+			else:
+				deleterious_steps += 1
+				
 			curr_id = df.loc[curr_id, "parent_id"]
 
 		paths.append(",".join(path))
@@ -86,6 +103,9 @@ def main():
 		temp["rolling_mean_500_volatility"] = fits.rolling(500).mean().var()
 		temp["rolling_mean_100_volatility"] = fits.rolling(100).mean().var()
 		temp["rolling_mean_50_volatility"] = fits.rolling(50).mean().var()
+		temp["x_magnitude"] = x_magnitude
+		temp["y_magnitude"] = y_magnitude
+		temp["total_magnitude"] = total_magnitude
 
 		all_data = pd.concat([all_data, temp])
 

### 2018-04-02 paper.tex
Update on Overleaf.

@@
 \usepackage{natbib,alifeconf}  %% The order is important
+
+% *****************
+%  TODOs:
+% *****************
+% [ ] Writing: Abstract
+% [ ] Writing: Results & discussion
+% [ ] Writing: Conclusions
+% [ ] Big 'ol citation pass: need to work appropriate citations into paper
 
@@
 
-\title{Something about lineage metrics}
-%\author{Chris Adami$^{1}$, Ryoichi Seki$^{1,2}$ \and Robel Yirdaw$^2$ \\
-%\mbox{}\\
-%$^1$California Institute of Technology, Pasadena, CA 91125 \\
-%$^2$California State University, Northridge, CA 91330 \\
-%adami@caltech.edu} % email of corresponding author
+\title{Understanding fine temporal scale evolutionary dynamics with lineage metrics}
+\author{Emily Dolson$^{1,2,3}$, Alexander Lalejini$^{1,2, 3}$, Steven Jorgensen$^{1,2}$ \and Charles Ofria$^{1,2, 3}$ \\
+\mbox{}\\
+$^1$BEACON Center for the Study of Evolution in Action, Michigan State University, East Lansing, MI, 48824 \\
+$^2$Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, 48824 \\
+$^3$Ecology, Evolutionary Biology, and Behavior Program, Michigan State University, East Lansing, MI, 48824 \\
+dolsonem@msu.edu} % email of corresponding author
 
@@
 Evolution is an emergent effect of events (\textit{e.g.} replication, variation, and competition) that occur on a much finer-grained temporal scale.
-While this is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results.
+While evolution's emergent nature is a large part of why it is so fascinating, it also presents challenges to studying the short-term mechanisms that govern long-term results.
 In computational evolutionary systems, we theoretically have adequate data to untangle these mechanisms.
 In practice, though, gathering appropriate data to do so can be challenging.
-Moreover, because of the vast difference between the scales that the causes and effects occur on, the quantity of such data can be overwhelming.
+Moreover, because of the vast difference between the scales that evolutionary causes and effects can occur on, the quantity of such data can be overwhelming.
 Both of these problems can be mitigated with a standardized set of diagnostic summary metrics that paint a picture of shorter-term evolutionary dynamics within a population.
-Here, we introduce a suite of such metrics that operate on lineages and provide an intuition for what they can tell us about evolution.
+Here, we suggest a suite of such metrics that operate on lineages and phylogenies.
+We then use experimental results to provide an intuition for what they can tell us about evolution.
 
@@
 A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation.
-Indeed, lineage analyses are a powerful tool for disentangling the evolution of complexity in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in modern wet lab experiments, allowing us to replay the tape of life in exacting detail to tease apart the evolutionary recipe for whatever complex feature we happen to be interested in.
+Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for fine-grained lineage tracking at resolutions that are impossible in modern wet lab experiments, allowing us to replay the tape of life in exacting detail and to tease apart the evolutionary recipe for whatever complex feature or phenomenon we happen to be interested in.
 Yet, tracking the full details of a single lineage, much less a population of lineages, can be computationally expensive and will inevitably generate an unwieldy amount of data that can be challenging to visualize and interpret [CITATIONS].
-Summary statistics can help alleviate these issues by focusing on general trends(?) for a population rather than for each individual.
 % @AML: Need a transition from here to next paragraph about how diagnostic/summary statistics can help
 % @SPJ: Something like this?
+%@ELD That looks like a good start, but we still need the part where we  actually discuss our methodology for creating these metrics. I'm going to try splitting it off into its own paragraph
 
-Here, we demonstrate three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility.
-For each, we discuss their application and our intuitions for what they can tell us about evolution.
+Summary statistics can help alleviate these issues by enabling the user to focus on aggregate trends across a population rather than needing to examine each individual.
+The question is how to effectively summarize what is essentially a path through fitness space.
+A useful abstraction is to treat the path as a sequence of states.
+Here, we will use phenotypes and genotypes as the states in the sequence, but we could just as easily some other descriptor of the lineage's position in the fitness landscape at a given point in time.
+% @ELD: Should we cite Alex's phenotypic plasticity stuff here?
+@AML: Nah, this paragraph is pretty sleek in making its points.
+I think the metrics section might be more appropriate.
+We can use it as an example.
+With this abstraction in hand, a few metrics are easily formalized: the number of unique states, the number of transitions between states, and the amount of time spent in each state. 
+Additionally, we may care about how the transitions between states happened.
+What mutations lead to them?
+Were those mutations beneficial, deleterious, or neutral at the time?
+Here, we choose a subset of these metrics that we expect will be particularly useful to explore further.
+
+Whereas a lineage represents the evolutionary history that lead up to a single individual, a phylogeny represents the evolutionary history of an entire population.
+Measurements that summarize phylogenies can provide useful insight into population-level evolutionary dynamics, such as diversification and co-existence among different clades.
+A variety of useful phylogeny measurements have already been developed by biologists~\citep{tucker_guide_2017}.
+These measurements tend to treat the phylogeny as a graph and make calculations about its topology.
+Tucker \textit{et al.} group them into three broad categories: assessments of the quantity of evolutionary history represented by a population, assessments of the amount of divergence within that evolutionary history, and assessments of the topological regularity of the phylogenetic tree.
+Such measurements can help quantify the behavior of the population as a whole, providing insight into interactions between its members.
+Thus, they are useful indicators of the presence of various types of eco-evolutionary dynamics.
+% @AML: I think the next bit (just moved it) might belong in the metrics section. Instead, we might want a little more elaboration on each of the categories of metrics.
+
+Here, we present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility.
+Additionally, we present four phylogeny measurements: most-recent common ancestor depth, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
+For each metric, we discuss its application and our intuitions for what it can tell us about evolution.
 We evaluate our intuitions on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths.
-% @AML: we'll want to say something here about how we talk about how these metrics extend to genetically discrete organisms (e.g. GP)
+For this work, we restrict our attention to asexually reproducing populations; however, we make suggestions as to how these metrics can be extended to sexual populations.
 
 \section{Metrics}
+
+\subsection{Lineage Metrics}
 % Outlining some points to (maybe) make here; probably too bogged down in details; not even sure we want to talk about lineages as graphs:
@@
 %  - For each, we discuss our intuition for what they can tell us (alone and in combination) about evolution. 
+Each of the three lineage metrics we discuss --- lineage length, mutation accumulation, and phenotypic volatility --- reduces a lineage to a linear sequence of states where each state represents an individual or a continuous sequence of individuals that share some common genotypic or phenotypic characteristic of interest.
+While we limit our focus to three lineage metrics, this abstraction places lineages in a form suitable for a wide range of measurements, including the direct application of many data mining techniques designed to operate over sequences.
+% @AML: TODO - will want to come up with a few examples/citations for such data mining techniques (can refer to data mining class notes/textbook).
 
-\subsection{Lineage Length}
+% @AML: this while paragraph could use some sanity checking/smoothing out. 
+Only lineages consisting of asexually reproducing ancestors where genetic material is exclusively vertically transmitted can be directly abstracted as a \textit{linear sequence} of states. 
+Sexual reproduction (and any form of horizontal gene transfer) complicates matters significantly as such lineages are more appropriately represented by trees rooted at the extant organism, branching for each contributor of genetic material.
+For each metric we present here, we limit our discussion to asexual populations; however, we suggest two approaches for generalizing to sexual populations: apply a lossy compression to reduce sexual lineages into linear sequences or extend our application of the metrics to operate over non-linear (tree) structured lineages.
+Sexual lineages can be compressed into linear sequences of states by modeling sexual reproduction events as asexual reproduction events, designating one parent to be a part of the lineage and considering the genetic contributions of other parents as sources of genetic variation (mutations).
+As previously stated, this approach is lossy, discarding potentially important parentage information.
+Alternatively, we can extend our metrics to operate over the more complex state sequences that constitute the lineages of sexually reproducing organisms.
+One such approach would be to consider all possible ancestor paths for an extant individual, taking an average of a measurement over each path.
+
+\subsubsection{Lineage Length}
 % Define/explanation
@@
 
-% Here, we...
+Lineage length describes the number of ancestral states traversed by a lineage.
+If a state represents a single individual along a lineage, lineage length is a measure of the number of generations on a lineage.
+This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined but instead determined by the life history strategies of organisms.
+In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates (\textit{e.g.} an r replicator strategy) and shorter lineages imply slower replication rates (\textit{e.g.} a k replicator strategy).
 
-\subsection{Mutation Accumulation}
+Lineage length becomes an increasingly flexible and informative metric if we begin to consider more abstract definitions of an ancestral state along a lineage.
+For example, we might measure the length of a lineage according to a particular phenotypic trait or genetic marker where an ancestral state represents a temporally continuous sequence of individuals that share the characteristics of interest.
+
+\subsubsection{Mutation Accumulation}
 Mutation Accumulation defines a set of measurements that track mutational changes across a lineage.
@@
 The type of a mutation can also be tracked, and can provide insight into the distribution of mutation types for a given lineage.
+Measures of mutation accumulation along the lineages of successful extant individuals can help to tease apart the relative importance of different types of mutational events when compared to what is expected by chance.
+
 In conjunction with collected fitness information, the class of a mutation (\textit{e.g.} beneficial, deleterious, or neutral) can also be tracked.
-% @SPJ: TODO
+Different evolutionary conditions are expected to cause different distributions of mutations along a lineage [CITE barrick and lenski genome dynamics]; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.
+The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape as well as how well a lineage was able to cross fitness valleys.
+An elevated measure of neutral mutations relative to beneficial or deleterious mutations might indicate that a lineage is spending most of its time drifting in a neutral fitness landscape.
+
 % Mutation accumulation over a lineage 
@@
 
-% Here, we...
+\subsubsection{Phenotypic Volatility}
 
-\subsection{Phenotypic Volatility}
+Phenotypic volatility addresses the rate at which the lineage changes phenotype.
+In systems with discrete/categorical phenotypes, this can be measured by summing the number of times the phenotype changes.
+A related but subtly different measurement in such systems is the number of unique phenotypes on a lineage.
+In most cases, these values will be very similar.
+A discrepancy between them, however, would be a clear signal of interesting dynamics.
+Specifically, it would suggest that the lineage were in a portion of the landscape where it could rapidly switch among different phenotypes.
+Such behavior could be indicative of some form of evolutionary bet-hedging.
+
+In systems with continuous-valued phenotypes, a subtly different approach is needed to measure phenotypic volatility. 
+% TODO
+
 % Define/describe
@@
 
-\subsection{Summary statistics}
+\subsubsection{Summary statistics}
 
@@
 There are two main approaches to doing so: 1) choose a small number of representative lineages from a given time point, 2) collect summary statistics about the distribution of metric values across the population.
+
+A single representative lineage can be chosen by selecting the lineage of the dominant organism (either the most fit or the more numerous; here we use the most fit).
+In populations with a diversity of coexisting strategies this approach can be uninformative as the dominant lineage is not necessarily representative of all successful lineages.
+An alternative in these scenarios is to filter out lineages that do not have offspring some predetermined number of generations later as such lineages were likely not representative of an important subset of the population.
+Still, any approach based on measuring only a subset of lineages can be challenging to interpret when the current dominant lineage (or lineages) is replaced with a different one; such changes can introduce a discontinuity if the value is being measured over time.
+If graceful responses to changes in which lineage is dominant are required, it can be advantageous to instead measure summary statistics (e.g.
+mean, variance, and range) across the entire population.
+
+In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be very similar to the average lineage in the population, as most of the population will be closely related.
+When the population contains more phylogenetic diversity, however, it is possible for the dominant lineage to differ from the mean.
+The nature of such differences is likely informative about the evolutionary dynamics occurring in the population.
+Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper.
+Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population.
+
+\subsection{Phylogeny metrics}
+
+These metrics operate on entire phylogenies rather than single lineages within a population.
+Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics.
+Because they factor in the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage.
+On the other hand, because most lineages in the population share a substantial amount of history, phylogeny metrics can often be calculated more rapidly than lineage metrics can be calculated across an entire population.
+Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc.
+Thus, when we refer to items in a phylogeny, we will use the term \textit{taxa}.
+
+A standard technique for saving memory and time when working with lineages in computational systems is to "prune" them, removing any taxa that are not either currently alive or the ancestor of taxa that are alive.
+Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies.
+Thus, for the remainder of this paper, we will assume we are only working with pruned phylogenies.
+
+In populations without ecological forces promoting coexistence, phylogenies should coalesce periodically, resulting in pruned lineages that mostly consist of a single path.
+When there is strong selection, this coalescence should happen even more rapidly.
+Thus, phylogenies with topologies that deviate from that expectation are an indication of ecological interactions within the population.
+The metrics discussed here can provide insight into the nature of those interactions and their long-term evolutionary effects.
+As a result, they are often referred to as phylogenetic diversity metrics~\citep{tucker_guide_2017}.
+
+An important distinction between phylogenies in natural vs computational systems to be aware of is that in natural systems phylogenies are generally inferred from extant taxa, whereas in computational systems they are directly measured.
+The inferred phylogenies used in natural systems do not contain internal nodes unless those nodes are branch points.
+They also do not contain history prior to the most recent common ancestor (MRCA) of all organisms in the extant population.
+For consistency, we will exclude pre-MRCA taxa from our analyses.
+However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.
+
+Here we provide a high-level summary of phylogeny metrics that we expect will be particularly useful.
+For more metrics and more detail on all of these metrics, see~\citep{winter_phylogenetic_2013, tucker_guide_2017}.
+
+\subsubsection{Depth of Most-Recent Common Ancestor}
+
+The depth of the MRCA (i.e.
+how many steps from the beginning of time did it occur?) is a useful metric because of the relatively high ratio of informativeness to ease of calculation.
+A more recent MRCA implies more frequent selective sweeps and less long-term stable coexistence between clades.
+The depth of the MRCA is so easily calculated because most any implementation of phylogeny pruning requires keeping track of it already.
+Measuring the frequency with which the MRCA changes (i.e.
+the number of coalescence events) can also by informative, as some conditions can inflate the length of the lineage relative to other conditions, without actually increasing the frequency of selective sweeps.
+This scenario is particularly likely when the population size is changing over time.
+A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never see a change in the MRCA after the very beginning of evolution.
+
+\subsubsection{Phylogenetic Richness}
+
+Measurements of phylogenetic richness quantify the total amount of evolutionary history that is contained in a set of taxa.
+The most traditional metric of phylogenetic richness is the one simply titled "Phylogenetic Diversity", which is calculated as the number of the nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}.
+Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}.
+A third approach is to calculate evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, and sum it across all extant taxa~\citep{tucker_guide_2017}.
+
+\subsubsection{Phylogenetic Divergence}
+
+Measurements of phylogenetic divergence quantify how distinct the taxa in the population are from each other, and are often the means of values that can be calculated for individual taxa.
+For instance, one option is to calculate the mean of the pairwise distances between all taxa in the population~\citep{webb_exploring_2000}.
+Similarly, phylogenetic divergence can also be calculated by calculating the evolutionary distinctiveness of each taxon in the population (as alluded to in the previous section) and taking the average of these values.
+
+\subsubsection{Phylogenetic Regularity}
+
+Measurements of phylogenetic regularity quantify how balanced the branches of the phylogeny are, and are often the variances of values that can be calculated for individual taxa.
+Just as the mean of the pairwise distances between all taxa in the population is a measurement of phylogenetic divergence, taking their variance is a measurement of phylogenetic regularity.
+The same is true of the variance of evolutionary distinctiveness across the population.
 
@@
 \section{Data Analysis}
-
+\subsection{3D visualizations}
 % Trim dramatically
@@
 
+\subsection{Metric analysis}
+
+We analyzed trends in our metrics using the R Statistical Computing Language~\citep{r_core_team_r:_2017}.
+Specifically, we used the ggplot2 library for all graphs included in this paper~\citep{wickham_ggplot2:_2009}.
+All analysis scripts are available at https://github.com/stevenjson/ALife2018-Lineage.
+
 \section{Results and Discussion}
@@
 \begin{figure}
-\includegraphics[width=3.5in]{figs/dom_deleterious_mutation_rate.png}
+\includegraphics[width=3.5in]{figs/mean_deleterious_mutation.png}
 \end{figure}
 \begin{figure}
-\includegraphics[width=3.5in]{figs/variance_deleterious_mutation_rate.png}
+\includegraphics[width=3.5in]{figs/mrca_depth_mutation.png}
 \end{figure}
+
+% \begin{figure}
+% \includegraphics[width=3.5in]{figs/variance_deleterious_mutation_rate.png}
+% \end{figure}
 
@@
 % ...
+
+\begin{figure}
+\includegraphics[width=3.5in]{figs/mean_deleterious_ts.png}
+\end{figure}
+\begin{figure}
+\includegraphics[width=3.5in]{figs/mrca_ts.png}
+\end{figure}
 

### 2018-04-02 analysis/data_analysis.Rmd
Update graphing script

--- 2018-03-28 analysis/data_analysis.Rmd
@@
 library(readr)
-
+library(reshape2)
+library(dplyr)
+library(scales)
 ```
@@
 real_value_data$seed <- as.factor(real_value_data$seed)
-real_value_data$ts <- as.factor(real_value_data$ts)
+#real_value_data$ts <- as.factor(real_value_data$ts)
 real_value_data$elite <- as.factor(real_value_data$elite)
-real_value_data$mutation_rate <- as.factor(real_value_data$mutation_rate)
+#real_value_data$mutation_rate <- as.factor(real_value_data$mutation_rate)
 tournament_data <- subset(real_value_data, real_value_data$selection == "Tournament" & real_value_data$elite == 0)
+melted_tournament_data <- melt(tournament_data, id.vars = c("update", "selection", "seed", "elite", "mutation_rate", "le", "p", "problem", "g", "ts"))
+end_points_vars_of_interest <- melted_tournament_data %>% filter(update==5000) %>% filter(variable %in% c("mrca_depth", "dominant_deleterious_steps", "mean_evolutionary_distinctiveness", "variance_evolutionary_distinctiveness", "current_phylogenetic_diversity", "mean_pairwise_distance", "variance_pairwise_distance", "ave_depth"))
+
+labels <- c(current_phylogenetic_diversity = "Phylogenetic diversity", dominant_deleterious_steps = "Dominant deleterious steps", mean_evolutionary_distinctiveness="Mean evo.
+distinctiveness", mean_pairwise_distance="Mean pairwise distance", mrca_depth="MRCA depth", variance_evolutionary_distinctiveness="Variance evo.
+distinctiveness", variance_pairwise_distance = "Variance pairwise distance")
+
 ```
@@
 ```{r}
-#see how many finished
-summary(subset(real_value_data, real_value_data$update == 5000))
+ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$mutation_rate==.001), aes(x=ts, y=value, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="line", fun.y=median)+ theme_classic() + scale_x_log10("Tournament size", breaks=c(2,4,8,16)) + scale_y_continuous("", labels=comma) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels)) + scale_color_discrete("Problem") + theme(legend.position = "bottom")
+ggsave("../figs/all_ts.png", width=7, height=5) 
+```
+
+```{r}
+ggplot(subset(end_points_vars_of_interest, end_points_vars_of_interest$ts==4), aes(x=mutation_rate, y=value, color=problem)) + stat_summary(fun.y = median, fun.ymin = function(x){quantile(x)[2]}, fun.ymax=function(x){quantile(x)[4]}) +stat_summary(geom="line", fun.y=median)+ theme_classic() + scale_x_log10("Mutation rate", labels=scientific) + scale_y_continuous("", labels=comma) + facet_wrap(~variable, scales="free_y", labeller = labeller(variable=labels)) +scale_color_discrete("Problem") +theme(legend.position = "bottom")
+ggsave("../figs/all_mutation_rate.png", width=7, height=5)
 
 ```
-
-First, let's make sure there's nothing weird going on with fitness:
-
-```{r, echo=FALSE}
-ggplot(tournament_data) + stat_summary(aes(x=update, y=max_fitness, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
-ggsave("../figs/all_tournament_fitness.png")
-```
-
-The most important thing to observe from that is probably that mutation rate seems to have a notable effect on the fitness dynamics of the Shubert function.
-
-Onto the metrics.
-
-## Phenotypic volatility
-
-```{r, echo=FALSE}
-ggplot(tournament_data) + stat_summary(aes(x=update, y=mean_phenotypic_volatility, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
-ggsave("../figs/all_phenotypic_volatility.png")
-```
-
-Oh right, that one kind of breaks in these problems...
-
-## Phenotype count
-
-```{r, echo=FALSE}
-ggplot(tournament_data) + stat_summary(aes(x=update, y=mean_unique_phenotypes, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
-ggsave("../figs/all_phenotype_count.png")
-
-```
-
-These too.
-
-## Deleterious steps
-
-```{r, echo=FALSE}
-ggplot(tournament_data) + stat_summary(aes(x=update, y=mean_deleterious_steps, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
-ggsave("../figs/all_deleterious_steps.png")
-
-```
-
-In general, weaker selection allows more deleterious steps, as does higher mutation rate.
-
-```{r, echo=FALSE}
-ggplot(tournament_data) + stat_summary(aes(x=update, y=dominant_deleterious_steps, color=ts, fill=ts), fun.data="mean_cl_boot", geom="smooth") + theme_classic() +facet_wrap(~problem*mutation_rate, ncol=8, scales="free")
-ggsave("../figs/dominant_deleterious_steps.png")
-
-```
-
-Looks pretty similar.
-
-```{r, echo=FALSE}
-ggplot(subset(tournament_data, tournament_data$update==5000)) + geom_point(aes(x=max_deleterious_steps, y=dominant_deleterious_steps, color=ts)) + facet_wrap(~problem*mutation_rate, ncol=8, scales="free") + theme_classic()
-```
-
-The relationship between max deleterious steps and dominant deleterious steps stops being a straight line at higher mutation rates.
-
-# Graphs for paper
-
-## Varying mutation rate:
-
-### Deleterious steps:
-
-```{r, echo=FALSE}
-ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=max_pairwise_distance, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
-```
-
-```{r, echo=FALSE}
-ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=dominant_deleterious_steps, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
-ggsave("../figs/dom_deleterious_mutation_rate.png")
-```
-
-There's possibly an interesting story to tell about variance:
-
-```{r, echo=FALSE}
-ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=variance_deleterious_steps, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
-ggsave("../figs/variance_deleterious_mutation_rate.png")
-```
-
-### Mutation magnitude
-
-```{r, echo=FALSE}
-ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4)) + geom_boxplot(aes(x=mutation_rate, y=mean_normal_mutations_on_lineage, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
-```
-
-Presumably the reason that .1 is ludircously high in everything except camel-back is that on the other landscapes it's jumping between peaks. Let's look at it without .1:
-
-```{r, echo=FALSE}
-ggplot(subset(tournament_data, tournament_data$update==5000 & tournament_data$ts == 4 & as.numeric(tournament_data$mutation_rate) < 8)) + geom_boxplot(aes(x=mutation_rate, y=mean_normal_mutations_on_lineage, color=mutation_rate)) + facet_wrap(~problem, ncol=8, scales="free") + theme_classic()
-```

### 2018-04-02 analysis/cec_python_library/extract_dominant_lineaage_variance.py

@@ -1,0 +1,125 @@
+#!/home/emily/anaconda2/bin/python
+###############################################################################
+# Version: 1.1
+# Last modified on: 3 April, 2016 
+# Developers: Michael G. Epitropakis
+#      email: m_(DOT)_epitropakis_(AT)_lancaster_(DOT)_ac_(DOT)_uk 
+###############################################################################
+from cec2013.cec2013 import *
+import numpy as np
+import sys
+import pandas as pd
+import glob
+import os
+
+def main():
+
+	glob_pattern = sys.argv[1]
+	outfilename = sys.argv[2]
+
+	all_data = pd.DataFrame()
+
+	# print(glob_pattern, glob.glob(glob_pattern))
+
+	problem_map = {0:4, 1:5, 2:6, 6:12}
+
+	for dirname in glob.glob(glob_pattern):
+		print(dirname)
+		run_log = dirname + "/run.log"
+		filename = dirname + "/pop_5000/phylogeny_5000.csv"
+
+		if not (os.path.exists(run_log) and os.path.exists(filename)):
+			print("skipped")
+			continue
+
+		local_data = {}
+
+		with open(run_log) as run_log_file:
+			for line in run_log_file:
+				if line.startswith("Doing initial"):
+					break
+				elif not line.startswith("set"):
+					continue
+				line = line.split()
+				local_data[line[1]] = line[2]
+
+		# Create function
+		f = CEC2013(problem_map[int(local_data["PROBLEM"])])
+		assert(f.get_dimension() == 2)			
+
+		df = pd.read_csv(filename, index_col="id")
+
+		df.rename(str.strip, axis="columns", inplace=True)
+		# print(df.columns.values)
+		genotypes = [i.strip("[] ").split() for i in df["info"]]
+		genotypes = map(lambda x: [float(x[0]), float(x[1])], genotypes)
+		df["fitness"] = [f.evaluate(np.array(x)) for x in genotypes]
+		genotypes = zip(*genotypes)
+
+		df["x"] = genotypes[0]
+		df["y"] = genotypes[1]
+
+		paths = []
+		ids  = []
+
+		next_gen = set([])
+		curr_id = int(df[df["num_orgs"] > 0]["fitness"].idxmax())
+
+		path = []
+		fits = []
[…]

### 2018-04-02 analysis/cec_python_library/extract_dominant_lineaage_variance.py
Added latest version of lineage variance

@@ -77,7 +77,7 @@ def main():
 			# print(df.loc[curr_id, :])
 			fits.append(df.loc[curr_id, "fitness"])
 			
-			path.append(" ".join(["{:.2f}".format(i) for i in [df.loc[curr_id, "x"], df.loc[curr_id, "y"], df.loc[curr_id, "fitness"]]]))
+			# path.append(" ".join(["{:.2f}".format(i) for i in [df.loc[curr_id, "x"], df.loc[curr_id, "y"], df.loc[curr_id, "fitness"]]]))
 			if (df.loc[curr_id, "parent_id"] == 1):
 				break
 			x_magnitude += abs(df.loc[df.loc[curr_id, "parent_id"], "x"] - df.loc[curr_id, "x"]) 
